---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
ds <- read.csv("activity.csv", na.strings = "NA")
ds[,"date"] <- lubridate::ymd(ds$date)
ds[,"steps"] <- as.numeric(ds[,"steps"])
```


## What is mean total number of steps taken per day?
```{r, message=FALSE}
library(dplyr)
daily <- ds %>% group_by(date) %>% summarize(total_steps = sum(steps,na.rm = TRUE))
hist(daily$total_steps, xlab = "daily steps", main = "histogram of the total number of steps taken each day", col = "azure")
```
  
The mean and median of the total number of steps taken per day are as follow    
```{r, message=FALSE, }
mean(daily$total_steps, na.rm = TRUE)
median(daily$total_steps, na.rm = TRUE)
```
  
## What is the average daily activity pattern?
```{r, message=FALSE}
daily_int <- ds %>% group_by(interval) %>% summarize(avg_steps = mean(steps,na.rm = TRUE))
plot(daily_int$interval,daily_int$avg_steps,type = "l", xlab = "5-min intervals", ylab = "average daily steps")
```

The 5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps is
```{r}
daily_int$interval[daily_int$avg_steps == max(daily_int$avg_steps)]
```
  
## Imputing missing values
The total number of missing values in the dataset (i.e. the total number of rows with NAs) is
```{r}
sum(is.na(ds$steps))
```
  
New dataset **ds1** is created which is equal in size to the original dataset (ds) but with the missing data filled in with the mean for that 5-minute interval.
```{r, cache=TRUE}
ds1 <-data.frame()
for(i in 1:nrow(ds)){
        int <-ds[i,"interval"]
        if (is.na(ds[i,"steps"])){
                ds1<-rbind(ds1, data.frame(steps = daily_int$avg_steps[daily_int$interval== int], date = ds[i,"date"], interval = int))
        }else {ds1<-rbind(ds1,data.frame(steps = ds[i,"steps"], date = ds[i,"date"], interval = int))}
}
str(ds1)
```
Histogram and mean/median calculations with the new dataset **ds1**
```{r}
daily1 <- ds1 %>% group_by(date) %>% summarize(total_steps = sum(steps,na.rm = TRUE))
hist(daily1$total_steps, xlab = "daily steps", main = "histogram of the total number of steps taken each day", col = "azure")
```

```{r, message=FALSE, }
mean(daily1$total_steps, na.rm = TRUE)
median(daily1$total_steps, na.rm = TRUE)
```

The impact of imputing missing data on the estimates of the total daily number of steps is shown below
```{r, message=TRUE}
par(mfrow = c(1,2))
hist(daily$total_steps, xlab = "daily steps", main = "original dataset", col = "azure", ylim = c(0,35))
hist(daily1$total_steps, xlab = "daily steps", main = "NA filled dataset", col = "azure")

t <- data.frame(mean = c(mean(daily$total_steps, na.rm = TRUE),mean(daily1$total_steps, na.rm = TRUE)), median = c(median(daily$total_steps, na.rm = TRUE),median(daily1$total_steps, na.rm = TRUE)), row.names = c("original dataset","NA filled dataset"))
print(t)
```
## Are there differences in activity patterns between weekdays and weekends?
New factor variable is created in the dataset (ds1) with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
ds1 <- ds1 %>% mutate(day = ifelse((weekdays(date) == "Saturday"|weekdays(date)== "Sunday"),"weekend","weekday"))
table(ds1$day)
```

```{r, message=FALSE}
daily_int1 <- ds1 %>% group_by(interval, day) %>% summarize(avg_steps = mean(steps,na.rm = TRUE))
ggplot2::qplot(interval,avg_steps,data = daily_int1, facets = day ~.,geom = "line", xlab = "5-min intervals", ylab = "average daily steps")
```